---
title: "Introduction to genr: design, track, and discover"
output:
  rmarkdown::html_document:
    self_contained: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Introduction to genr: design, track, and discover}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = TRUE)
```

Gen is a sequence version control system built around a graph database.
Rather than storing a sequence as a flat string, Gen breaks it into
nodes, each holding a short subsequence, connected by edges into a
graph. A path through that graph reconstructs a full sequence. When a
change is recorded, Gen adds new nodes and edges to represent the
altered subsequence and the new junctions around it. A path can then
take an alternative route through the new nodes, or skip nodes entirely
in the case of a deletion. Changes can be naturally occurring or
deliberately engineered. What makes this especially powerful is that the
graph does not just capture changes over time: it can represent the full
set of expected or observed variants across haplotypes, design libraries,
or a pangenome, all coexisting in one database with their differences
always queryable.

This vignette walks through a concrete lab scenario that shows where
that design pays off: you design an edit, apply it, confirm it with
search, export for sequencing, and then discover an unintended mutation
in the result.

## Setup

Install and load the package, then initialise a workspace and open a
repository handle.

```{r setup, message=FALSE, warning=FALSE}
library(genr)

# Create a fresh workspace directory and open a repository handle.
ws <- file.path(tempdir(), "puc19_demo")
dir.create(ws)
setwd(ws)

repo <- Repository(NULL)
```

`Repository()` discovers the `.gen/` directory from the current working
directory (or from the `path` argument) and opens the graph and
operations databases. All subsequent operations go through this object.

## Import an annotated vector

We start with a fully annotated pUC19 sequence from NEB. The GenBank
format carries feature annotations (genes, promoters, the MCS) that will
be preserved in the graph and displayed in the viewer.

```{r import}
gbk_path <- system.file("extdata", "puc19.gbk", package = "genr")
sample_wt <- repo$import_genbank(gbk_path, sample = "wt")
sg_wt <- sample_wt$block_groups[[1]]
cat("Sequence graph:", sg_wt$name(), "| sample:", sg_wt$sample_name(), "\n")
```

## Visualise the reference

```{r plot-wt}
p <- plot(sg_wt)

# Navigate to the MCS — annotated features are first-class, so you can
# jump directly to any named feature rather than specifying coordinates.
annotations <- sg_wt$list_annotations()
mcs <- Filter(function(a) grepl("MCS|misc_feature|multiple.cloning", a$name, ignore.case = TRUE), annotations)
if (length(mcs) > 0) p$go_to(mcs[[1]])

p
```

The viewer renders a static snapshot of the sequence graph. Annotations
from the GenBank file (lacZα, bla, the MCS) appear as coloured tracks
below the sequence. `sg_wt$list_annotations()` returns all named features in
the database; passing one to `go_to()` centres the viewport on that
feature with no coordinate look-up required.

For a fully interactive, pannable, zoomable view, run `gen view` from
the command line in the same workspace.

## Design a Golden Gate conversion

The pUC19 multiple cloning site (MCS, positions 396–452) contains a set
of traditional type-II restriction sites. We want to replace it with a
minimal Golden Gate entry site: two BsaI recognition sequences (GGTCTC)
flanking a stuffer, oriented to cut inward and expose compatible 4-nt
overhangs for assembly.

```{r design-gg}
# Original MCS (1-based 396..452):
#   GAATTCGAGCTCGGTACCCGGGGATCCTCTAGAGTCGACCTGCAGGCATGCAAGCTT
#   EcoRI SacI  KpnI  SmaI  BamHI XbaI    SalI   PstI SphI HindIII
#
# Replacement: two BsaI sites flanking a stuffer, retaining EcoRI/HindIII ends.
#   GAATTC [BsaI→ GGTCTCA] [AATG overhang] [stuffer] [GCAT overhang] [←BsaI TGAGACC] AAGCTT

golden_gate_mcs <- "GAATTCGGTCTCAAATGCATCATCATCATGCATTGAGACCAAGCTT"

sample_gg <- repo$update_with_sequence(
  sequence    = golden_gate_mcs,
  sample      = "wt",
  new_sample  = "gg_design",
  # MCS coordinates from the GenBank misc_feature annotation (1-based 396..452),
  # converted to 0-based half-open interval for the region string.
  # Hardcoded for now — a future repo$get_annotation_coordinates() will allow
  # looking these up by name directly from the database.
  region_name = "pUC19:395-452"
)
sg_gg <- sample_gg$block_groups[[1]]
```

`update_with_sequence` creates a new sample ("gg_design") that shares
the entire graph with "wt" except at the edited region, where the path
diverges through the new MCS nodes, and returns that sample's sequence
graphs directly.

## Inspect the design

### Confirm the BsaI sites with search

Before synthesis, we verify the Golden Gate sites landed correctly by
searching for the BsaI recognition sequence on both strands. `search()`
uses `sequence_kind = "dna"` by default, which automatically searches
the reverse complement as well as the forward sequence.

```{r search-bsai}
# GGTCTC is the BsaI recognition sequence (cuts 1 nt downstream on sense strand).
# DNA-mode search finds both the forward site and its reverse-complement counterpart.
hits <- repo$search("GGTCTC", c(sg_gg$id()), "dna")
cat("BsaI sites found:", length(hits[[1]]$matches), "\n")
```

```{r plot-gg}
p_gg <- plot(sg_gg)
p_gg$go_to(hits[[1]]$matches[[1]])
p_gg
```

The viewer centres on the first BsaI site, which sits at the boundary of
the new Golden Gate MCS. Both paths are visible: the wild-type MCS and
the new Golden Gate site diverge at position 395 and reconverge 57 bp
later (original) or 46 bp later (new design).

Three hits. To understand why, run the same search against the
wild-type:

```{r search-wt}
hits_wt <- repo$search("GGTCTC", c(sg_wt$id()), "dna")
wt_count <- if (length(hits_wt) > 0) length(hits_wt[[1]]$matches) else 0L
cat("BsaI sites in wild-type:", wt_count, "\n")   # 1
```

The wild-type already carries one BsaI site on the reverse strand at
position 1765, inside the **bla** coding sequence. The design added
exactly two new sites (3 − 1 = 2), so the flanking Golden Gate sites are
present and correctly oriented. The pre-existing site is the problem: if
left in place it will be cut alongside the intended assembly sites,
fragmenting the AmpR cassette. It needs to be silently mutated out
before the construct goes to synthesis.

The `search()` method also accepts degenerate IUPAC codes, so you can
query by recognition sequence directly without pre-computing all
variants. For example, EcoRI's sequence `GAATTC` is unambiguous, but a
site like HincII (`GTYRAC`, where Y = C/T and R = A/G) expands to four
possible hexamers; pass the IUPAC string and `search()` matches all of
them:

```{r search-iupac}
hits_hincii <- repo$search("GTYRAC", c(sg_gg$id()), "dna")
cat("HincII sites found:", length(hits_hincii[[1]]$matches), "\n")
```

For large genomes, consider calling `repo$build_index()` before
searching. The index speeds up exact queries; degenerate IUPAC patterns
always use a full scan regardless.

### Navigate and highlight search results

Each entry in the `matches` list is a locus record with `start`, `end`,
`blocks`, and `strand`. Pass a locus directly to `go_to_match` to move
the viewport to it (or `go_to_match(locus, center = TRUE)` to center
it instead), and to `highlight_match` to colour it on the canvas.

The convenience method `go_to()` accepts both search match loci and
annotation records returned by `list_annotations()`. Annotations from
one sample can be used to navigate a plot of a different sample. This
is safe and stable because annotations are stored as references to graph
nodes, not as linear sequence coordinates. Nodes are shared across
samples in the same repository, so an annotation created for the "wt"
sample points to exactly the same graph nodes that appear in the
"gg_design" or "sequenced" graphs.

```{r highlight-bsai}
matches <- hits[[1]]$matches

# Jump to the first BsaI site and highlight both sites.
p_gg$go_to_match(matches[[1]])
p_gg$highlight_match(matches[[1]], color = "cyan")
p_gg$highlight_match(matches[[2]], color = "cyan")
p_gg
```

`clear_highlights()` removes all highlights without affecting the
viewport:

```{r clear-highlights}
p_gg$clear_highlights()
p_gg
```

## Export and send for sequencing

```{r export}
repo$export_fasta("pUC19_gg_design.fa", sample = "gg_design")
```

You send `pUC19_gg_design.fa` to your synthesis provider. A few days
later sequencing results arrive as a VCF.

## Import the sequencing result and find a surprise

```{r update-vcf}
vcf_path <- system.file("extdata", "puc19_sequenced.vcf", package = "genr")

samples_seq <- repo$update_with_vcf(
  filename       = vcf_path,
  sample         = "sequenced",
  reference = c("gg_design")
)
sg_seq <- samples_seq[[1]]$block_groups[[1]]
```

`update_with_vcf()` returns a list of samples, one per output sample
name — here there's a single output sample ("sequenced"), so we take
its first (and only) sequence graph.

Plot the sequenced sample:

```{r plot-sequenced}
p_seq <- plot(sg_seq)
p_seq
```

The graph now shows a branch at position 1904, outside the MCS and inside
the **bla** (β-lactamase / AmpR) coding sequence. The sequencing run
picked up a C→T substitution that was not in the design. Because Gen
stores every sample as a path through the same graph, the mutation shows
up as a new node rather than a silent overwrite: the gg_design path and
the sequenced path diverge exactly here.

This is the kind of mutation that is easy to miss in a flat-file
workflow. Gen surfaces it because "what changed between two samples" is
a native database query, not a diff of exported files.

## Version control on the command line

Everything above runs inside R. Gen also ships a command-line interface
that adds the operations that make it a true version control system:

``` sh
gen init
gen import --fasta reference.fa --sample wt
gen checkout -b gg_design
gen apply --vcf site.vcf
gen push origin gg_design
gen pull
gen log
```

`push`, `pull`, `checkout`, `log`, and `merge` work like their git
equivalents but operate on sequence graphs rather than text files. A Gen
remote holds the full graph history: who edited which sample, when, and
how the paths diverged.

The R bindings and the CLI share the same database, so you can design a
construct in R, push it from the terminal, and a colleague can pull the
graph and visualise it in their own R session, with the full annotation
and variant history intact.

To learn more about the CLI, run `gen --help` or visit the project
documentation.

## Further capabilities

The sections below cover the full R API surface. All examples are
self-contained and assume `repo <- Repository(NULL)` in a workspace
initialised with `gen init` (or the R equivalent).

### Additional import formats

#### GFA

GFA (Graphical Fragment Assembly) files encode sequence graphs directly.
Import a `.gfa` file as a new sample:

```{r import-gfa, eval=FALSE}
repo$import_gfa("assembly.gfa", sample = "assembly")
```

Export any sample back to GFA. The optional `node_max` argument splits
nodes longer than the given length, useful for tools that expect bounded
node sizes:

```{r export-gfa, eval=FALSE}
repo$export_gfa("output.gfa", sample = "assembly", node_max = 1000L)
```

#### Reference FASTA

`import_reference_fasta()` differs from `import_fasta()` in that it
marks the imported sample as a reference in the database. This is the
right starting point when you plan to call variants relative to an
external aligner's reference: subsequent `update_with_vcf()` calls can
name this sample in the `reference` argument without ambiguity.

```{r import-ref-fasta, eval=FALSE}
repo$import_reference_fasta("hg38.fa", reference = "hg38")
```

Export any sample's annotations and sequence to GenBank format:

```{r export-genbank, eval=FALSE}
repo$export_genbank("pUC19_gg_design.gb", sample = "gg_design")
```

### Bioconductor integration

#### Named sequences and XStringSet

`import_bioconductor()` accepts a named character vector,
`DNAStringSet`, or any `XStringSet` and imports each entry as a block
group without writing an intermediate FASTA file. Pass `repo` to target
an existing workspace; omit it to use the current working directory.

```{r import-bioc, eval=FALSE}
library(Biostrings)

seqs <- DNAStringSet(c(
  plasmid_a = "ATCGATCGATCG",
  plasmid_b = "TTGCTTGCTTGC"
))
import_bioconductor(seqs, sample = "constructs", repo = repo)
```

#### Genomic regions (GRanges)

`import_granges()` imports a list of genomic intervals as sequence graphs.
Coordinates follow Bioconductor convention (1-based, closed start-end)
and are converted to 0-based half-open storage internally. Each unique
reference sequence is stored once as a shared node, so overlapping
regions over the same chromosome do not duplicate sequence data.

```{r import-granges, eval=FALSE}
library(GenomicRanges)
library(Biostrings)

genome <- DNAStringSet(c(
  chr1 = "ACGTACGTACGTACGT",
  chr2 = "TTGCTTGCTTGCTTGC"
))

gr <- GRanges(
  seqnames = c("chr1", "chr1", "chr2"),
  ranges   = IRanges(start = c(1, 9, 5), end = c(8, 16, 12))
)
names(gr) <- c("gene_a", "gene_b", "gene_c")

import_granges(gr, genome, sample = "hg38", repo = repo)
```

### Combinatorial library design

`import_library()` builds a combinatorial sequence graph from a list of
named part vectors. Each element of `parts_list` is one column of the
design: a named character vector (or `DNAStringSet`) where each name is
a part label and each value its sequence. Parts are stored as shared
nodes in a compact graph; `search()` screens every junction
simultaneously without enumerating individual combinations.

The example below designs an expression cassette with three promoters,
three RBS variants, and two CDSes (18 unique combinations):

```{r import-library, eval=FALSE}
upstream <- c(upstream = "AATTCGGATCCAAGCTT")

promoters <- c(
  pTrc = "TTGACAATTAATCATCCGGCTCGTATAATGTGTGG",
  pT7  = "TAATACGACTCACTATA",
  pLac = "AATTGTGAGCGGATAACAATT"
)

rbs_variants <- c(
  rbs_strong = "AAAGAGGAGAAA",
  rbs_medium = "AAGAGGAG",
  rbs_weak   = "AGGAG"
)

cds_variants <- c(
  gfp = "ATGAGTAAAGGAGAAGAACTTTTCACTGG",
  rfp = "ATGGCTTCCTCCGAAGACGTTATCAAAGAG"
)

terminator <- c(terminator_T1 = "GCGCAACGCAATTAATGTGAGTTAGCTCACTCATTAGGCACCCCAGGC")

cassette_sg <- repo$import_library(
  library_name = "expression-cassette",
  parts_list   = list(upstream, promoters, rbs_variants, cds_variants, terminator),
  sample       = "design-v1"
)

plot(cassette_sg)
```

`DNAStringSet` columns work identically to named character vectors and
can be mixed freely in the same `parts_list`. For parts defined by
genomic coordinates, use `import_library_from_granges()` with a
`seq_containers` list; see the `library_from_granges.R` example in
`inst/examples/` for the full pattern.

#### Applying a library update to an existing sample

`update_with_library()` works like `update_with_sequence()` but replaces
a region with a set of combinatorial variants. The `path_name` argument
selects the region using the same `"<name>:<start>-<end>"` format as
other update functions:

```{r update-library, eval=FALSE}
repo$update_with_library(
  sample          = "gg_design",
  new_sample_name = "gg_tagged",
  path_name       = "pUC19:1200-1270",
  parts_list      = list(
    c(tag_a = "ATGATGATG", tag_b = "TGATGATGA")
  )
)
```

### Graph partitioning

#### Subgraphs

`derive_subgraph()` creates a new sequence graph that is a coordinate slice
of an existing one. The `region` argument uses `"<name>:<start>-<end>"`
(0-based half-open) or just `"<name>"` for the full sequence:

```{r derive-subgraph, eval=FALSE}
mcs_region_sg <- repo$derive_subgraph(
  sample     = "gg_design",
  new_sample = "mcs_region",
  region     = "pUC19:390-460"
)
```

Sequence graphs expose `subgraph()` as a shorthand that returns the new
sequence graph immediately:

```{r bg-subgraph, eval=FALSE}
mcs_sg <- sg_gg$subgraph("mcs_v2", 390L, 460L, NULL)
plot(mcs_sg)
```

#### Chunks and stitching

`sg$chunks()` splits a sequence graph into contiguous fragments and returns
them as a list of sequence graphs. Supply `chunk_size` for uniform splits
or `breakpoints` (an integer vector of positions) for explicit cut
points:

```{r derive-chunks, eval=FALSE}
chunks <- sg_gg$chunks("gg_500bp_chunks", integer(0), 500L, NULL)
cat("Chunks created:", length(chunks), "\n")
```

`stitch()` reverses the operation, concatenating a list of
same-sample sequence graphs end-to-end into a new sequence graph:

```{r stitch, eval=FALSE}
reassembled <- stitch(
  repo,
  bgs        = chunks,
  new_sample = "gg_reassembled",
  new_region = "pUC19.reassembled"
)
```

`repo$derive_chunks()` is the lower-level counterpart that creates
chunks in-place and returns the resulting sample directly.

### Annotation tracks

Any `gen_plot` can display additional annotation panels below the
sequence graph. Add a GFF3 or BED file as a named track; the `sample`
argument selects which sample's path defines the coordinate space:

```{r add-track-file, eval=FALSE}
p_gg$add_track_file("features.gff3", name = "Features", sample = "gg_design")
p_gg
```

Annotation groups stored in the database (created automatically on
GenBank import) can be added by group name:

```{r add-track-group, eval=FALSE}
p_gg$add_track_group("GenBank annotations")
p_gg
```
