---
title: "Introduction to genr: design, track, and discover"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to genr: design, track, and discover}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

Gen is a sequence version control system built around a graph database.  Rather
than storing sequences as flat strings, Gen stores them as a sequence graph where
every sample is a named path through that graph.  Edits branch the graph rather
than overwriting it, so every version of a sequence coexists in the same
database and their differences are always queryable.

This vignette walks through a concrete lab scenario that shows where that
design pays off: you design an edit, apply it, confirm it with search, export
for sequencing — and then discover an unintended mutation in the result.

## Setup

Install and load the package, then initialise a workspace and open a
repository handle.

```{r setup}
library(genr)

# Create a fresh workspace directory and open a repository handle.
ws <- file.path(tempdir(), "puc19_demo")
dir.create(ws)
setwd(ws)

repo <- Repository()
```

`Repository()` discovers the `.gen/` directory from the current working
directory (or from the `path` argument) and opens the graph and operations
databases.  All subsequent operations go through this object.

## Import an annotated vector

We start with a fully annotated pUC19 sequence from NEB.  The GenBank format
carries feature annotations (genes, promoters, the MCS) that will be preserved
in the graph and displayed in the viewer.

```{r import}
gbk_path <- system.file("extdata", "puc19.gbk", package = "genr")
repo$import_genbank(gbk_path, sample = "wt")
```

List the block groups in the repository to confirm the import:

```{r list-bgs}
bgs <- repo$get_block_groups()
bg_wt <- bgs[[1]]
cat("Block group:", bg_wt$name, "| sample:", bg_wt$sample_name, "\n")
```

## Visualise the reference

```{r plot-wt}
p <- repo$plot(bg_wt)
p
```

The viewer renders a pannable, zoomable canvas of the sequence graph.  At this
stage the graph for the wild-type sample is a single linear path.  Annotations
from the GenBank file (lacZα, bla, the MCS) appear as tracks above the
sequence.

## Design a Golden Gate conversion

The pUC19 multiple cloning site (MCS, positions 396–452) contains a set of
traditional type-II restriction sites.  We want to replace it with a minimal
Golden Gate entry site — two BsaI recognition sequences (GGTCTC) flanking a
stuffer, oriented to cut inward and expose compatible 4-nt overhangs for
assembly.

```{r design-gg}
# Original MCS (1-based 396..452):
#   GAATTCGAGCTCGGTACCCGGGGATCCTCTAGAGTCGACCTGCAGGCATGCAAGCTT
#   EcoRI SacI  KpnI  SmaI  BamHI XbaI    SalI   PstI SphI HindIII
#
# Replacement: two BsaI sites flanking a stuffer, retaining EcoRI/HindIII ends.
#   GAATTC [BsaI→ GGTCTCA] [AATG overhang] [stuffer] [GCAT overhang] [←BsaI TGAGACC] AAGCTT

golden_gate_mcs <- "GAATTCGGTCTCAAATGCATCATCATCATGCATTGAGACCAAGCTT"

repo$update_with_sequence(
  sequence    = golden_gate_mcs,
  sample      = "wt",
  new_sample  = "gg_design",
  # MCS coordinates from the GenBank misc_feature annotation (1-based 396..452),
  # converted to 0-based half-open interval encoded in the region string.
  # Programmatic annotation lookup is not yet exposed in the R API;
  # a future repo$get_annotations(bg) call would return these directly.
  region_name = "pUC19:395-452"
)
```

`update_with_sequence` creates a new sample ("gg_design") that shares the
entire graph with "wt" except at the edited region, where the path diverges
through the new MCS nodes.

## Inspect the design

```{r get-gg-bg}
bg_gg <- repo$get_block_groups_by_collection(bg_wt$collection)
bg_gg <- Filter(function(x) x$sample_name == "gg_design", bg_gg)[[1]]
```

```{r plot-gg}
p_gg <- repo$plot(bg_gg)
p_gg
```

Zoom into the MCS region to see both paths: the wild-type MCS and the new
Golden Gate site diverge at position 395 and reconverge 57 bp later (original)
or 46 bp later (new design).

### Confirm the BsaI sites with search

Before synthesis, we verify the Golden Gate sites landed correctly by searching
for the BsaI recognition sequence on both strands.

```{r build-index}
repo$build_index(bgs = list(bg_gg))
```

```{r search-bsai}
# GGTCTC is the BsaI recognition sequence (cuts 1 nt downstream on sense strand).
# Searching both the sense strand and its reverse complement locates both sites.
hits <- repo$search("GGTCTC", bgs = list(bg_gg))
cat("BsaI sites found:", length(hits[[1]]$matches), "\n")
```

Two hits — one forward, one reverse — confirm the orientation of the flanking
BsaI sites.  As a sanity check, the same search against the wild-type sample
should return zero hits:

```{r search-wt}
repo$build_index(bgs = list(bg_wt))
hits_wt <- repo$search("GGTCTC", bgs = list(bg_wt))
cat("BsaI sites in wild-type:", length(hits_wt[[1]]$matches), "\n")   # 0
```

### Navigate and highlight search results

Each entry in the `matches` list is a locus record with `start`, `end`,
`blocks`, and `strand`.  Pass a locus directly to `goto_match` to centre the
viewport on it, and to `highlight_match` to colour it on the canvas.

```{r highlight-bsai}
matches <- hits[[1]]$matches

# Jump to the first BsaI site and highlight both sites.
p_gg$goto_match(matches[[1]])
p_gg$highlight_match(matches[[1]], color = "cyan")
p_gg$highlight_match(matches[[2]], color = "cyan")
p_gg
```

`clear_highlights()` removes all highlights without affecting the viewport:

```{r clear-highlights}
p_gg$clear_highlights()
p_gg
```

## Export and send for sequencing

```{r export}
repo$export_fasta("pUC19_gg_design.fa", sample = "gg_design")
```

You send `pUC19_gg_design.fa` to your synthesis provider.  A few days later
sequencing results arrive as a VCF.

## Import the sequencing result — and find a surprise

```{r update-vcf}
vcf_path <- system.file("extdata", "puc19_sequenced.vcf", package = "genr")

repo$update_with_vcf(
  filename       = vcf_path,
  sample         = "sequenced",
  reference = c("gg_design")
)
```

Plot the sequenced sample:

```{r get-sequenced-bg}
bg_seq <- repo$get_block_groups_by_collection(bg_wt$collection)
bg_seq <- Filter(function(x) x$sample_name == "sequenced", bg_seq)[[1]]
```

```{r plot-sequenced}
p_seq <- repo$plot(bg_seq)
p_seq
```

The graph now shows a branch at position 1904 — outside the MCS, inside the
**bla** (β-lactamase / AmpR) coding sequence.  The sequencing run picked up a
C→T substitution that was not in the design.  Because Gen stores every sample
as a path through the same graph, the mutation shows up as a new node rather
than a silent overwrite: the gg_design path and the sequenced path diverge
exactly here.

This is the kind of mutation that is easy to miss in a flat-file workflow.
Gen surfaces it because "what changed between two samples" is a native
database query, not a diff of exported files.

## Version control on the command line

Everything above runs inside R.  Gen also ships a command-line interface that
adds the operations that make it a true version control system:

```sh
gen init
gen import --fasta reference.fa --sample wt
gen checkout -b gg_design
gen apply --vcf site.vcf
gen push origin gg_design
gen pull
gen log
```

`push`, `pull`, `checkout`, `log`, and `merge` work like their git equivalents
but operate on sequence graphs rather than text files.  A Gen remote holds the
full graph history: who edited which sample, when, and how the paths diverged.

The R bindings and the CLI share the same database, so you can design a
construct in R, push it from the terminal, and a colleague can pull the graph
and visualise it in their own R session — with the full annotation and variant
history intact.

To learn more about the CLI, run `gen --help` or visit the project documentation.
