---
title: "Introduction to genr: design, track, and discover"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Introduction to genr: design, track, and discover}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = TRUE)
```

Gen is a sequence version control system built around a graph database.
Rather than storing sequences as flat strings, Gen stores them as a
sequence graph where every sample is a named path through that graph.
Edits branch the graph rather than overwriting it, so every version of a
sequence coexists in the same database and their differences are always
queryable.

This vignette walks through a concrete lab scenario that shows where
that design pays off: you design an edit, apply it, confirm it with
search, export for sequencing — and then discover an unintended mutation
in the result.

## Setup

Install and load the package, then initialise a workspace and open a
repository handle.

```{r setup}
library(genr)

# Create a fresh workspace directory and open a repository handle.
ws <- file.path(tempdir(), "puc19_demo")
dir.create(ws)
setwd(ws)

repo <- Repository()
```

`Repository()` discovers the `.gen/` directory from the current working
directory (or from the `path` argument) and opens the graph and
operations databases. All subsequent operations go through this object.

## Import an annotated vector

We start with a fully annotated pUC19 sequence from NEB. The GenBank
format carries feature annotations (genes, promoters, the MCS) that will
be preserved in the graph and displayed in the viewer.

```{r import}
gbk_path <- system.file("extdata", "puc19.gbk", package = "genr")
repo$import_genbank(gbk_path, sample = "wt")
```

List the block groups in the repository to confirm the import:

```{r list-bgs}
bgs <- repo$get_block_groups()
bg_wt <- bgs[[1]]
cat("Block group:", bg_wt$name, "| sample:", bg_wt$sample_name, "\n")
```

## Visualise the reference

```{r plot-wt}
p <- repo$plot(bg_wt)
p
```

The viewer renders a pannable, zoomable canvas of the sequence graph. At
this stage the graph for the wild-type sample is a single linear path.
Annotations from the GenBank file (lacZα, bla, the MCS) appear as tracks
above the sequence.

## Design a Golden Gate conversion

The pUC19 multiple cloning site (MCS, positions 396–452) contains a set
of traditional type-II restriction sites. We want to replace it with a
minimal Golden Gate entry site — two BsaI recognition sequences (GGTCTC)
flanking a stuffer, oriented to cut inward and expose compatible 4-nt
overhangs for assembly.

```{r design-gg}
# Original MCS (1-based 396..452):
#   GAATTCGAGCTCGGTACCCGGGGATCCTCTAGAGTCGACCTGCAGGCATGCAAGCTT
#   EcoRI SacI  KpnI  SmaI  BamHI XbaI    SalI   PstI SphI HindIII
#
# Replacement: two BsaI sites flanking a stuffer, retaining EcoRI/HindIII ends.
#   GAATTC [BsaI→ GGTCTCA] [AATG overhang] [stuffer] [GCAT overhang] [←BsaI TGAGACC] AAGCTT

golden_gate_mcs <- "GAATTCGGTCTCAAATGCATCATCATCATGCATTGAGACCAAGCTT"

repo$update_with_sequence(
  sequence    = golden_gate_mcs,
  sample      = "wt",
  new_sample  = "gg_design",
  # MCS coordinates from the GenBank misc_feature annotation (1-based 396..452),
  # converted to 0-based half-open interval for the region string.
  # Hardcoded for now — a future repo$get_annotation_coordinates() will allow
  # looking these up by name directly from the database.
  region_name = "pUC19:395-452"
)
```

`update_with_sequence` creates a new sample ("gg_design") that shares
the entire graph with "wt" except at the edited region, where the path
diverges through the new MCS nodes.

## Inspect the design

```{r get-gg-bg}
bg_gg <- Filter(function(x) x$sample_name == "gg_design", repo$get_block_groups())[[1]]
```

```{r plot-gg}
p_gg <- repo$plot(bg_gg)
p_gg
```

Zoom into the MCS region to see both paths: the wild-type MCS and the
new Golden Gate site diverge at position 395 and reconverge 57 bp later
(original) or 46 bp later (new design).

### Confirm the BsaI sites with search

Before synthesis, we verify the Golden Gate sites landed correctly by
searching for the BsaI recognition sequence on both strands. `search()`
uses `sequence_kind = "dna"` by default, which automatically searches
the reverse complement as well as the forward sequence.

```{r search-bsai}
# GGTCTC is the BsaI recognition sequence (cuts 1 nt downstream on sense strand).
# DNA-mode search finds both the forward site and its reverse-complement counterpart.
hits <- repo$search("GGTCTC", bgs = list(bg_gg))
cat("BsaI sites found:", length(hits[[1]]$matches), "\n")
```

Three hits. To understand why, run the same search against the wild-type:

```{r search-wt}
hits_wt <- repo$search("GGTCTC", bgs = list(bg_wt))
wt_count <- if (length(hits_wt) > 0) length(hits_wt[[1]]$matches) else 0L
cat("BsaI sites in wild-type:", wt_count, "\n")   # 1
```

The wild-type already carries one BsaI site — on the reverse strand at
position 1765, inside the **bla** coding sequence. The design added
exactly two new sites (3 − 1 = 2), so the flanking Golden Gate sites
are present and correctly oriented. The pre-existing site is the
problem: if left in place it will be cut alongside the intended assembly
sites, fragmenting the AmpR cassette. It needs to be silently mutated
out before the construct goes to synthesis.

The `search()` method also accepts degenerate IUPAC codes, so you can
query by recognition sequence directly without pre-computing all
variants. For example, EcoRI's sequence `GAATTC` is unambiguous, but a
site like HincII (`GTYRAC`, where Y = C/T and R = A/G) expands to four
possible hexamers — pass the IUPAC string and `search()` matches all of
them:

```{r search-iupac}
hits_hincii <- repo$search("GTYRAC", bgs = list(bg_gg))
cat("HincII sites found:", length(hits_hincii[[1]]$matches), "\n")
```

For large genomes, consider calling `repo$build_index()` before
searching. The index speeds up exact queries; degenerate IUPAC patterns
always use a full scan regardless.

### Navigate and highlight search results

Each entry in the `matches` list is a locus record with `start`, `end`,
`blocks`, and `strand`. Pass a locus directly to `goto_match` to centre
the viewport on it, and to `highlight_match` to colour it on the canvas.

```{r highlight-bsai}
matches <- hits[[1]]$matches

# Jump to the first BsaI site and highlight both sites.
p_gg$goto_match(matches[[1]])
p_gg$highlight_match(matches[[1]], color = "cyan")
p_gg$highlight_match(matches[[2]], color = "cyan")
p_gg
```

`clear_highlights()` removes all highlights without affecting the
viewport:

```{r clear-highlights}
p_gg$clear_highlights()
p_gg
```

## Export and send for sequencing

```{r export}
repo$export_fasta("pUC19_gg_design.fa", sample = "gg_design")
```

You send `pUC19_gg_design.fa` to your synthesis provider. A few days
later sequencing results arrive as a VCF.

## Import the sequencing result — and find a surprise

```{r update-vcf}
vcf_path <- system.file("extdata", "puc19_sequenced.vcf", package = "genr")

repo$update_with_vcf(
  filename       = vcf_path,
  sample         = "sequenced",
  reference = c("gg_design")
)
```

Plot the sequenced sample:

```{r get-sequenced-bg}
bg_seq <- Filter(function(x) x$sample_name == "sequenced", repo$get_block_groups())[[1]]
```

```{r plot-sequenced}
p_seq <- repo$plot(bg_seq)
p_seq
```

The graph now shows a branch at position 1904 — outside the MCS, inside
the **bla** (β-lactamase / AmpR) coding sequence. The sequencing run
picked up a C→T substitution that was not in the design. Because Gen
stores every sample as a path through the same graph, the mutation shows
up as a new node rather than a silent overwrite: the gg_design path and
the sequenced path diverge exactly here.

This is the kind of mutation that is easy to miss in a flat-file
workflow. Gen surfaces it because "what changed between two samples" is
a native database query, not a diff of exported files.

## Version control on the command line

Everything above runs inside R. Gen also ships a command-line interface
that adds the operations that make it a true version control system:

``` sh
gen init
gen import --fasta reference.fa --sample wt
gen checkout -b gg_design
gen apply --vcf site.vcf
gen push origin gg_design
gen pull
gen log
```

`push`, `pull`, `checkout`, `log`, and `merge` work like their git
equivalents but operate on sequence graphs rather than text files. A Gen
remote holds the full graph history: who edited which sample, when, and
how the paths diverged.

The R bindings and the CLI share the same database, so you can design a
construct in R, push it from the terminal, and a colleague can pull the
graph and visualise it in their own R session — with the full annotation
and variant history intact.

To learn more about the CLI, run `gen --help` or visit the project
documentation.

## Further capabilities

The sections below cover the full R API surface. All examples are
self-contained and assume `repo <- Repository()` in a workspace
initialised with `gen init` (or the R equivalent).

### Additional import formats

#### GFA

GFA (Graphical Fragment Assembly) files encode sequence graphs directly.
Import a `.gfa` file as a new sample:

```{r import-gfa, eval=FALSE}
repo$import_gfa("assembly.gfa", sample = "assembly")
```

Export any sample back to GFA. The optional `node_max` argument splits
nodes longer than the given length, useful for tools that expect bounded
node sizes:

```{r export-gfa, eval=FALSE}
repo$export_gfa("output.gfa", sample = "assembly", node_max = 1000L)
```

#### Reference FASTA

`import_reference_fasta()` differs from `import_fasta()` in that it
marks the imported sample as a *reference* in the database. This is the
right starting point when you plan to call variants relative to an
external aligner's reference: subsequent `update_with_vcf()` calls can
name this sample in the `reference` argument without ambiguity.

```{r import-ref-fasta, eval=FALSE}
repo$import_reference_fasta("hg38.fa", reference = "hg38")
```

#### GenBank updates and export

`update_with_genbank()` applies sequence edits and annotation changes
encoded in a GenBank file to an existing sample. Set `create_missing =
TRUE` to auto-create block groups for features not yet in the database:

```{r update-genbank, eval=FALSE}
repo$update_with_genbank(
  "corrected_annotation.gb",
  sample         = "wt",
  create_missing = TRUE
)
```

Export any sample's annotations and sequence to GenBank format:

```{r export-genbank, eval=FALSE}
repo$export_genbank("pUC19_gg_design.gb", sample = "gg_design")
```

### Bioconductor integration

#### Named sequences and XStringSet

`import_bioconductor()` accepts a named character vector,
`DNAStringSet`, or any `XStringSet` and imports each entry as a block
group without writing an intermediate FASTA file. Pass `repo` to target
an existing workspace; omit it to use the current working directory.

```{r import-bioc, eval=FALSE}
library(Biostrings)

seqs <- DNAStringSet(c(
  plasmid_a = "ATCGATCGATCG",
  plasmid_b = "TTGCTTGCTTGC"
))
import_bioconductor(seqs, sample = "constructs", repo = repo)
```

#### Genomic regions (GRanges)

`import_granges()` imports a list of genomic intervals as block groups.
Coordinates follow Bioconductor convention (1-based, closed
\[start, end\]) and are converted to 0-based half-open storage
internally. Each unique reference sequence is stored once as a shared
node, so overlapping regions over the same chromosome do not duplicate
sequence data.

```{r import-granges, eval=FALSE}
library(GenomicRanges)
library(Biostrings)

genome <- DNAStringSet(c(
  chr1 = "ACGTACGTACGTACGT",
  chr2 = "TTGCTTGCTTGCTTGC"
))

gr <- GRanges(
  seqnames = c("chr1", "chr1", "chr2"),
  ranges   = IRanges(start = c(1, 9, 5), end = c(8, 16, 12))
)
names(gr) <- c("gene_a", "gene_b", "gene_c")

import_granges(gr, genome, sample = "hg38", repo = repo)
```

### Combinatorial library design

`import_library()` builds a combinatorial block group from a list of
named part vectors. Each element of `parts_list` is one column of the
design: a named character vector (or `DNAStringSet`) where each name is
a part label and each value its sequence. Every combination of one part
per column becomes a path through the resulting graph.

The example below designs an expression cassette with three promoters,
three RBS variants, and two CDSes — 18 unique combinations:

```{r import-library, eval=FALSE}
upstream <- c(upstream = "AATTCGGATCCAAGCTT")

promoters <- c(
  pTrc = "TTGACAATTAATCATCCGGCTCGTATAATGTGTGG",
  pT7  = "TAATACGACTCACTATA",
  pLac = "AATTGTGAGCGGATAACAATT"
)

rbs_variants <- c(
  rbs_strong = "AAAGAGGAGAAA",
  rbs_medium = "AAGAGGAG",
  rbs_weak   = "AGGAG"
)

cds_variants <- c(
  gfp = "ATGAGTAAAGGAGAAGAACTTTTCACTGG",
  rfp = "ATGGCTTCCTCCGAAGACGTTATCAAAGAG"
)

terminator <- c(terminator_T1 = "GCGCAACGCAATTAATGTGAGTTAGCTCACTCATTAGGCACCCCAGGC")

repo$import_library(
  library_name = "expression-cassette",
  parts_list   = list(upstream, promoters, rbs_variants, cds_variants, terminator),
  sample       = "design-v1"
)

cassette_bgs <- Filter(function(x) x$name == "expression-cassette", repo$get_block_groups())
repo$plot(cassette_bgs[[1]])
```

`DNAStringSet` columns work identically to named character vectors and
can be mixed freely in the same `parts_list`. For parts defined by
genomic coordinates, pass `GRanges` columns alongside a `seq_containers`
list; see the `library_from_granges.R` example in `inst/examples/` for
the full pattern.

#### Applying a library update to an existing sample

`update_with_library()` works like `update_with_sequence()` but replaces
a region with a set of combinatorial variants. The `path_name` argument
selects the region using the same `"<name>:<start>-<end>"` format as
other update functions:

```{r update-library, eval=FALSE}
repo$update_with_library(
  sample          = "gg_design",
  new_sample_name = "gg_tagged",
  path_name       = "pUC19:1200-1270",
  parts_list      = list(
    c(tag_a = "ATGATGATG", tag_b = "TGATGATGA")
  )
)
```

### Graph partitioning

#### Subgraphs

`derive_subgraph()` creates a new block group that is a coordinate slice
of an existing one. The `region` argument uses `"<name>:<start>-<end>"`
(0-based half-open) or just `"<name>"` for the full sequence:

```{r derive-subgraph, eval=FALSE}
repo$derive_subgraph(
  sample     = "gg_design",
  new_sample = "mcs_region",
  region     = "pUC19:390-460"
)
```

Block groups expose `subgraph()` as a shorthand that returns the new
block group immediately:

```{r bg-subgraph, eval=FALSE}
mcs_bg <- bg_gg$subgraph("mcs_v2", start = 390L, end = 460L)
repo$plot(mcs_bg)
```

#### Chunks and stitching

`bg$chunks()` splits a block group into contiguous fragments and returns
them as a list of block groups. Supply `chunk_size` for uniform splits
or `breakpoints` (an integer vector of positions) for explicit cut
points:

```{r derive-chunks, eval=FALSE}
chunks <- bg_gg$chunks("gg_500bp_chunks", chunk_size = 500L)
cat("Chunks created:", length(chunks), "\n")
```

`repo$stitch()` reverses the operation, concatenating a list of
same-sample block groups end-to-end into a new block group:

```{r stitch, eval=FALSE}
reassembled <- repo$stitch(
  bgs        = chunks,
  new_sample = "gg_reassembled",
  new_region = "pUC19.reassembled"
)
```

`repo$derive_chunks()` is the lower-level counterpart that creates
chunks in-place (returning a status string); retrieve the results
afterwards with `repo$get_block_groups()` filtered by name.

### Annotation tracks

Any `gen_plot` can display additional annotation panels below the
sequence graph. Add a GFF3 or BED file as a named track — the `sample`
argument selects which sample's path defines the coordinate space:

```{r add-track-file, eval=FALSE}
p_gg$add_track_file("features.gff3", name = "Features", sample = "gg_design")
p_gg
```

Annotation groups stored in the database (created automatically on
GenBank import) can be added by group name:

```{r add-track-group, eval=FALSE}
p_gg$add_track_group("GenBank annotations")
p_gg
```

`clear_tracks()` removes all track panels without affecting the viewport
or highlights:

```{r clear-tracks, eval=FALSE}
p_gg$clear_tracks()
p_gg
```
