---
title: "Combinatorial expression cassette library for carotenoid biosynthesis in yeast"
output:
  rmarkdown::html_document:
    self_contained: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Combinatorial expression cassette library for carotenoid biosynthesis in yeast}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = TRUE)
```

The Yeast Toolkit (YTK) and MoClo-yeast workflows use BsmBI (recognition
sequence `CGTCTC`) as their Golden Gate enzyme: each cassette is flanked
by outward-facing BsmBI sites that expose 4-nt overhangs for modular
ligation. If an internal BsmBI site exists anywhere in the assembled
construct, the enzyme will cut it during the Golden Gate reaction and
shatter the cassette, hence it's important to use parts that have been
'domesticated' so they are free from restriction sites.

One issue that is harder to prevent is the junction-emergent site:
neither part carries BsmBI on its own, but when two specific parts are
placed in sequence the recognition site appears at their boundary.
One way to prevent this is by generating all possible combinations to
screen one at a time, which can quickly go exponential. We do not have
to do that, however, and can use `search()` directly on the sequence
graph and find matches across any combination of junctions.

**Scenario:** Expressing the bacterial geranylgeranyl pyrophosphate (GGPP)
synthase crtE, the committed step in carotenoid biosynthesis, in
Saccharomyces cerevisiae as part of a lycopene production strain. Three
constitutive promoters and two terminators drawn from the YTK part
collection are combined with two crtE variants (native and
codon-optimised) in a 3 × 2 × 2 factorial design of 12 expression
cassettes. During cloning, you notice that some combinations appear to
assemble much more efficiently than others.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(genr)
library(Biostrings)
library(S4Vectors)

ws <- file.path(tempdir(), "yeast_carotenoid")
dir.create(ws, showWarnings = FALSE)
setwd(ws)

repo <- Repository(NULL)
```

## Define parts

Parts are `DNAStringSet` objects. The `mcols()` metadata carries a `role`
column (promoter, cds, terminator, or scar) that will colour each
annotation group in the viewer.

Backbone scars are single-part columns so every path through the graph
is a complete, self-contained expression cassette, ready for Golden Gate
ligation into the pYTK001 destination vector.

```{r define-parts}
# Helper: build a DNAStringSet with a role annotation column.
make_parts <- function(nms, seqs, role) {
  dss <- DNAStringSet(setNames(seqs, nms))
  mcols(dss)$role <- role
  dss
}

# Upstream backbone scar, the left BsmBI assembly site.
left_scar <- make_parts(
  "left_scar",
  "AACTAGTATGGCGCGCCAACAAGGATCTGCGG",
  "scar"
)

# Three constitutive promoters from the YTK collection.
# TEF1p (strong) ends with ...GCGTCT, the first six bases of the BsmBI
# recognition sequence.  When followed by a CDS that begins with C, the
# junction spells CGTCTC and the cassette is destroyed during assembly.
promoters <- make_parts(
  c("TEF1p", "PGK1p", "HXT7p"),
  c(
    # TEF1p: strong, ~150 bp abbreviated.  Ends with GCGTCT.
    paste0(
      "AGATCTGTTGACAAATAATCATCCGGCTCGTAATGTGTGGAATTGTGAGCGGATAACAAT",
      "TTCACCAGTTCCCAACCTTACTCGGCACAAATCACCTTAATTTGATATTTATATCAAAGA",
      "ATTTCTTTTGTATGGCGTCT"
    ),
    # PGK1p: medium strength constitutive promoter.
    paste0(
      "AGTACTTTTTTTTTTATTTTTTTTTTTTTTTGAATTTTTTTTATTTTTTTCAATTCAATT",
      "CATCATTTTTTTTTTTTTTTTTTTTTTTCTTCTTCTTTCTTTCTTTTTTTTTATTTTTTT",
      "CATTTTCTTTTTTTTTTCAATT"
    ),
    # HXT7p: strong under glucose limitation (useful for fed-batch).
    paste0(
      "ATCAGTGAGCAGTATGATGGTTATATTATATAGAGCATATATAGACATATATTATATTAA",
      "ATAGATATAATTATATAATATAAGAAATAATAATATAAGAAATAATAATATAAGAAATAAT",
      "ATCAGTGAGCAACTAGCAC"
    )
  ),
  "promoter"
)

# Two crtE variants (GGPP synthase from Erwinia uredovora).
# crtE_native: unoptimised bacterial codons, ORF starts with CATG.
#   TEF1p + crtE_native junction: ...GCGTCT | CATG... = CGTCTC = BsmBI site!
# crtE_coopt: S. cerevisiae codon-optimised, ORF starts with ATGG.
#   TEF1p + crtE_coopt junction: ...GCGTCT | ATGG... = CGTCTA = no BsmBI site.
cds_variants <- make_parts(
  c("crtE_native", "crtE_coopt"),
  c(
    paste0(
      "CATGGTTACTATCCCAGAAAACGTTGGTGCGAACATGGGCATCGTCGAAGCTGATGACGA",
      "CGCTGCGGAATCCCAGTCTCAAGCTGCAATTTCCCAAATCAGACAAGCAATGATTTCTGA",
      "AGGTATTCACCAATGCAAGAAATGA"
    ),
    paste0(
      "ATGGTTACCTCTCCAGAAAACGTTGGTGCTAACATGGGTATCGTCGAAGCTGACGACGAC",
      "GCAGCTGAATCCCAATCTCAAGCAGCGATTTCCCAAATCAGACAAGCGATGATTTCTGAA",
      "GGTATCCATCAATGTAAGAAATGA"
    )
  ),
  "cds"
)

# Two terminators from the YTK collection.
terminators <- make_parts(
  c("CYC1t", "ADH1t"),
  c(
    paste0(
      "CATGTAATTAGTTATGTCACGCTTACATTCACGCCCTCCCCCACATCCGCTCTAACCAAA",
      "AATGGCCCAAAACTCGCCCCATTTCACGTTTATTATTGATTTTTTTTTTTTTTTTTCTTT",
      "TTTTTTTTTTTTTTTTTTTTTTT"
    ),
    paste0(
      "GCGAATTTCTTATGATTTATGATTTTTATTATTTATTATTTTTTATTATCTTTTTATTATT",
      "TTATTATTTATAATTTAATATATAAATAATATATTTAAATATATATTTTTTTATTATTTTAT",
      "TATTTATAATTTAATATATAAATAAT"
    )
  ),
  "terminator"
)

# Downstream backbone scar, the right BsmBI assembly site.
right_scar <- make_parts(
  "right_scar",
  "GCGGCCGCAAGCTTAGATCTCGCGGCCATACT",
  "scar"
)

parts_list <- list(left_scar, promoters, cds_variants, terminators, right_scar)
n_combos <- prod(lengths(parts_list))
cat(paste(lengths(parts_list), collapse = " × "), "=", n_combos, "combinations\n")
```

## Import the library

`import_library()` accepts a list of `DNAStringSet` columns and builds a
compact sequence graph where parts are shared nodes. Search traverses the
graph directly, screening every junction simultaneously without enumerating
individual constructs.

```{r import-library}
repo$import_library("lycopene-cassette", parts_list)

sg <- Filter(function(x) x$name() == "lycopene-cassette", repo$get_sequence_graphs())[[1]]
cat("Sequence graph:", sg$name(), "\n")
cat("Sample:        ", sg$sample_name(), "\n")
```

## Visualise the library graph

The annotation group added automatically on import labels every part.
Because `GenPlot` calls `auto_load_annotation_groups()` at construction
time, the track appears without an explicit `add_track_group()` call.

The `colors` parameter accepts a function that receives each annotation
and returns a hex color string.  Here we map the `role` field stored in
`ann$metadata` (propagated from `mcols()$role` on import) to a fixed palette.

```{r role-colors}
ROLE_COLORS <- c(
  scar       = "#aaaaaa",
  promoter   = "#4393c3",
  cds        = "#1a9850",
  terminator = "#e08214"
)
```

```{r plot-library}
p <- plot(sg, rows = 20L,
          colors = function(ann) ROLE_COLORS[[ann$metadata$role]])
p
```

The fork-and-rejoin topology of the 12-path graph is immediately visible:
three promoter branches merge at the CDS column, split again into two CDS
variants, reconverge at the terminator column, and so on.

## Build the design matrix

```{r design-matrix}
design <- expand.grid(
  promoter   = names(promoters),
  cds        = names(cds_variants),
  terminator = names(terminators),
  stringsAsFactors = FALSE
)
design$construct_id <- seq_len(nrow(design))
design
```

## Screen the full library for BsmBI

A single `search()` traverses all 12 paths simultaneously, including
every inter-part junction. BsmBI (`CGTCTC`) would cut the cassette during
the Golden Gate reaction; any internal site must be eliminated before
synthesis.

```{r search-bsmbi}
hits <- repo$search("CGTCTC", c(sg$id()), "dna")
n_sites <- length(hits[[1]]$matches)
cat("BsmBI sites found in the library graph:", n_sites, "\n")
```

The search returns unique site occurrences in the sequence graph, one
per distinct location in the shared graph structure. That one site sits
at the TEF1p→crtE_native junction and will be traversed by every
construct that pairs those two parts, regardless of which terminator
follows. Translating from graph site to blocked constructs happens in
the design matrix below.

## Diagnose: the site is junction-emergent

One site found, yet neither part contains BsmBI individually:

```{r diagnose}
count_bsmbi <- function(seq) {
  site <- "CGTCTC"
  n <- 0L
  for (i in seq_len(nchar(seq) - nchar(site) + 1L))
    if (substr(seq, i, i + nchar(site) - 1L) == site) n <- n + 1L
  n
}

all_parts <- c(promoters, cds_variants, terminators)
cat("BsmBI in individual parts:\n")
for (nm in names(all_parts)) {
  n <- count_bsmbi(as.character(all_parts[[nm]]))
  flag <- if (n > 0L) " ← !" else ""
  cat(sprintf("  %-18s %d site(s)%s\n", nm, n, flag))
}

tef1p_seq   <- as.character(promoters[["TEF1p"]])
native_seq  <- as.character(cds_variants[["crtE_native"]])
coopt_seq   <- as.character(cds_variants[["crtE_coopt"]])

cat("\nTEF1p ends:              ...", substr(tef1p_seq,  nchar(tef1p_seq)  - 11L, nchar(tef1p_seq)),  "\n")
cat("crtE_native starts:         ", substr(native_seq,  1L, 12L), "...\n")
cat("crtE_coopt starts:          ", substr(coopt_seq,   1L, 12L), "...\n")

junction_bad  <- paste0(tef1p_seq, native_seq)
junction_good <- paste0(tef1p_seq, coopt_seq)
cat("\nBsmBI at TEF1p + crtE_native junction:", count_bsmbi(junction_bad),  "\n")
cat("BsmBI at TEF1p + crtE_coopt junction: ", count_bsmbi(junction_good), "\n")
```

The recognition sequence `CGTCTC` is split across the junction:

- TEF1p ends `...GCGTCT` (the first six bases of the BsmBI octamer)
- crtE_native begins `CATG...` (supplying the terminal `C`)

Together they complete `CGTCTC`. This site exists on every path through
the graph that pairs TEF1p with crtE_native, regardless of which
terminator is chosen.

## Highlight the BsmBI site on the graph


```{r highlight-site}
hits <- repo$search("CGTCTC", c(sg$id()), "dna")
p2 <- plot(sg, rows=16L, cols=56L,
           colors = function(ann) ROLE_COLORS[[ann$metadata$role]])
m <- hits[[1]]$matches[[1]]
p2$go_to_match(m, center = TRUE)
p2$highlight_match(m, color = "red")

p2
```

The highlighted region straddles the TEF1p→crtE_native edge. Because
the site is stored once in the shared graph structure, a single highlight
marks all paths that would be cut: here, both the CYC1t and the ADH1t
variants of the TEF1p×crtE_native cassette.

## Flag blocked constructs

```{r flag-constructs}
design$blocked <- design$promoter == "TEF1p" & design$cds == "crtE_native"

blocked <- design[design$blocked, ]
clean   <- design[!design$blocked, ]

cat(nrow(blocked), "blocked construct(s):\n")
print(blocked[, c("construct_id", "promoter", "cds", "terminator")], row.names = FALSE)

cat("\n", nrow(clean), "clean constructs proceed to synthesis.\n")
```

The two blocked cassettes (TEF1p × crtE_native, regardless of terminator)
cannot be assembled by Golden Gate. The remaining ten are safe.

## Next steps

In practice, each part would be flanked by inward-facing BsmBI recognition
sequences plus enough additional handle sequence to seat PCR primers and
give the enzyme room to cut cleanly. The library graph holds the internal
part content; the flanking architecture is applied at the cloning design
stage.

Gen records every import and update as an operation. To attach the
screening conclusion to the `import_library` operation, open the
operations browser from the terminal:

``` sh
gen operations --interactive
```

Select the lycopene-cassette import entry to add a message, for example:
"Screening complete: TEF1p×crtE_native blocked by junction-emergent
BsmBI site at promoter-CDS boundary. Ten combinations cleared."

To redesign the library with a different set of parts, create a new
branch from the operation just before the import. `HEAD~1` is shorthand
for "one step back from the current operation" (analogous to `HEAD~1` in
git):

``` sh
gen checkout -b lycopene-v2 HEAD~1
```

Then modify the parts list in R and call `import_library()` again on the
new branch. The original branch and all its screening history remain
intact; you can pick up that experiment again at any point.
```
